Researchers preferentially collaborate with same-gendered colleagues across the life sciences

### 2018-06-06 Gender_homophily_analysis.Rmd
Tweak discussion

@@

### Summary output of the `brms` model used to derive the means in Figure 3
```{r}
cat(readLines('data/brms.output.txt'), sep = '\n')
```

>>>>

This figure illustrates that single-author papers make up a small minority of the total, for the majority of disciplines we examined. Indeed for many disciplines, the majority of papers have at least five authors.

@@

### Summary output of the `brms` model used to derive the means in Figure 3
Note that the 95% credible intervals for the population-level intercept do not overlap zero, which indicates "significant" positive $\alpha'$ among 2-author papers. Furthermore, $\alpha'$ is significantly higher for 3-, 4-, and 5-plus-author papers than for 2-author papers. The R^2 of the model is 0.44, and the random effects appear to explain non-trivial amounts of variation.
```{r}
cat(readLines('data/brms.output.txt'), sep = '\n')
```

### 2018-06-06 manuscript/Manuscript_Claire.Rmd
Acknowledgements
Tweak discussion

# Discussion
We found evidence that researchers preferentially publish with same-gendered coauthors, even after implementing stringent controls for Wahlund effects (\autoref{wahlund_plot}). Our study therefore reaffirms earlier studies' conclusions [e.g. @Ferber_1980; @McDowell_1992; @Bentley_2003; @Crow_2015; @Ghiasi_2015; @Araujo_2017a; @Fahmy_2017; @Jadidi_2017; @Teele_2017; @Zettler_2017] and establishes their generality across the life sciences. Relatively few journals had $\alpha'$ values below zero, and almost no journals showed statistically significant gender heterophily after controlling for multiple testing. The excess of same-gender coauthorships was quite large: many journals had $\alpha' > 0.1$, indicating that the gender ratio of men's and women's coauthors differs by >10% in absolute terms. In relative terms, our findings are even more striking: for example, if men have 20% female coauthors and women have 30% (i.e. $\alpha' = 0.1$ for a field with a male-biased gender ratio), then women publish with women 50% more often than men do.

An important limitation of our study is that we cannot reliably determine the cause(s) of the observed excess of same-gender coauthorships. As well as conscious or unconcious selection of collaborators based on gender, our results could be partly explained by uncontrolled Wahlund effects. However, we suspect the contribution of these to be minor, for four reasons: we found positive $\alpha'$ after controlling for three obvious sources of Wahlund effect; there was no inflation of $\alpha'$ in highly multidisciplinary journals; restricting the data by country yielded similar estimates of $\alpha'$; and we showed that differences in gender ratio between career stages are unlikely to fully explain our results. On balance, we believe the data provide good evidence that researchers preferentially select same-gendered collaborators, although the strength of this preference is uncertain.

@@

We also found that gender homophily was marginally stronger in 2015-2016 relative to 2005-2006. Although this trend might reflect a change in the gender preferences of researchers seeking collaborators over time, there are alternative (and perhaps more likely) explanations. For example, this trend might result from the increasing number of women working in senior positions in STEMM over the past decade [e.g. @Long_2015; @Bendels_2018; @McKenzie_2017]. As shown in \autoref{simulation}, if enough coauthorships are between junior and senior researchers, a large gender gap between career stages can give the appearance of heterophily. As this gender gap between career stages lessens, the observed values of $\alpha$ may increase.

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or sexism from men [@jagsi_2016], which would clearly be concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since Sheltzer and Smith [-@sheltzer_2014]'s results might reflect discrimination against women during hiring [@Moss_2012], or avoidance by women of elite research groups (e.g. due to 'imposter syndrome' or the perception that these groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality [which is contentious; @Garfield_2006], this result provides non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013].

Homophily might also have more benign causes. Collaboration is often most enjoyable and productive when working with like-minded people, who might be same-gendered more often than not. We also suppose that some people consciously choose to preferentially collaborate with women in order to help close the gender gap in the workforce; this would create homophily if women do this more than men. In support of this interpretation, women appear more likely than men to promote the work of female collagues by inviting them to give talks [@nittrouer_2018; @debarre_2018]. Related to this, Ghiasi et al. [-@Ghiasi_2015] concluded that women in engineering are "compliant [in reproducing] male-dominated scientific structures" because they do not collaborate often enough with other women; their data suggest that coauthorships between two women are about 30% more frequent than expected under random assortment (see their Figure 7). In contrast, we propose that men as well as women should ensure that they are not inadvertently overlooking or excluding female colleagues, particularly since men are disproportionately represented among senior researchers [@Shaw_2012].

@@

# Acknowledgements
CM was supported by the Academy of Finland (284666 to the Centre of Excellence in Biological Interactions).

>>>>

# Discussion
We found evidence that researchers preferentially publish with same-gendered coauthors, even after implementing stringent controls for Wahlund effects (\autoref{wahlund_plot}). Our study therefore reaffirms earlier studies' conclusions [e.g. @Ferber_1980; @McDowell_1992; @Bentley_2003; @Crow_2015; @Ghiasi_2015; @Araujo_2017a; @Fahmy_2017; @Jadidi_2017; @Teele_2017; @Zettler_2017] and establishes their generality across the life sciences. Relatively few journals had $\alpha'$ values below zero, and almost no journals showed statistically significant gender heterophily after controlling for multiple testing. The excess of same-gender coauthorships was quite large: many journals had $\alpha' > 0.1$, indicating that the gender ratio of men's and women's coauthors differs by >10% in absolute terms. In relative terms, our findings are even more striking: for example, if men have 20% female coauthors and women have 30% (i.e. $\alpha' = 0.1$ in a field with a male-biased gender ratio), then women publish with women 50% more often than men do.

An important limitation of our study is that we cannot reliably determine the cause(s) of the observed excess of same-gender coauthorships. As well as the obvious interpretation -- conscious or unconcious selection of same-gender collaborators by men, women, or both -- our results could be partly explained by uncontrolled Wahlund effects. However, we suspect the contribution of these to be minor, for four reasons: we found positive $\alpha'$ after controlling for three obvious sources of Wahlund effect; there was no inflation of $\alpha'$ in highly multidisciplinary journals; restricting the data by country yielded similar estimates of $\alpha'$; and we showed that differences in gender ratio between career stages are unlikely to fully explain our results. On balance, we believe the data suggest that it is likely that some researchers do preferentially select same-gendered collaborators, although the effect size of this preference is difficult to measure.

@@

We also found that gender homophily was marginally stronger in 2015-2016 relative to 2005-2006. Although this trend might reflect a change in the gender preferences of researchers seeking collaborators, there are alternative (and perhaps more likely) explanations. For example, this trend might result from the increasing number of women working in senior positions in STEMM over the past decade [e.g. @Long_2015; @Bendels_2018; @McKenzie_2017]. As shown in \autoref{simulation}, if enough coauthorships are between junior and senior researchers, a large gender gap between career stages can give the appearance of heterophily. As this gender gap between career stages lessens, the observed values of $\alpha'$ may increase.

Regarding our finding of weaker homophily among 2-author papers, we suspect that many 2-author teams comprise a student/postdoc and a senior staff member, making these teams especially likely to be mixed-gender, due to the elevated gender gap among senior researchers [@Shaw_2012; @Holman_2018]. Assuming this interpretation is correct, this result suggests that our overall $\alpha'$ may underestimate the strength of peoples' preferences for same-gendered collaborators; essentially, women seeking a senior collaborator could be constrained to work mostly with men, meaning that people's ideal and realised gender preferences would be mismatched. On a related note, Ghiasi et al. [-@Ghiasi_2015] argue that women in engineering are "compliant [in reproducing] male-dominated scientific structures" because they do not collaborate often enough with other women; for reference, their Figure 7 suggests that coauthorships between women are roughly 30% more frequent than expected under random assortment. By contrast, we feel that it is not helpful to recommend that women collaborate primarily with other women, e.g. because this constrains women's options and may be counter-productive (particularly in fields like engineering, where 90% of professors are men [@Shaw_2012]). Instead, we suggest that researchers of both genders can help to close the gender gap in science. In the context of collaboration, one way to do this is to undertake self-examination to ensure that one is not inadvertently overlooking or excluding female potential students and colleagues. One should also take care to treat male and female collaborators equally, e.g. in terms of training and mentoring, allocation of work, and how one frames or promotes the collaboration (e.g. in conference presentations or on a website); evidence suggests that unconscious bias causes people to undervalue women's research achievements [@Moss_2012], and to assign menial or under-valued tasks to women and more prestigious and employable tasks (e.g. writing and project leadership) to men [@macaluso_2016].

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment or sexism from men [@jagsi_2016], which would clearly be concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since Sheltzer and Smith [-@sheltzer_2014]'s results might reflect discrimination against women during hiring [@Moss_2012], or avoidance by women of elite research groups (e.g. due to 'imposter syndrome' or a perception that some groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality [which is contentious; @Garfield_2006], our results provide non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013]. Another issue is that if many collaborations are between established researchers, there will be an excess of male-male collaborations in fields where women in senior positions are rare; some of the observed homophily might therefore reflect the elevated gender gap among senior researchers.

On the other hand, homophily might have more benign causes. Collaboration is often most enjoyable and productive when working with like-minded people, who might be same-gendered more often than not. We also suppose that some people consciously choose to preferentially collaborate with women in order to help close the gender gap in the workforce; this would create homophily if women do this more than men. In support of this interpretation, women appear more likely than men to promote the work of female colleagues by inviting them to give talks [@nittrouer_2018; @debarre_2018]. Given that many collaborative research projects unfortunately involve a gendered division of labour [@macaluso_2016], working with a same-gendered colleague may provide exposure to new parts of the research process, and (especially for the minority gender) a welcome change of pace.

@@

# Acknowledgements
CM was supported by the Academy of Finland (284666 to the Centre of Excellence in Biological Interactions). We thank Devi Stuart-Fox and Dominique Potvin for helpful discussion.
